---
title: "temp"
author: "Matt"
date: '2023-05-24'
output: html_document
---




# Load Cellranger data into R

## Load cellranger matrix into matrix

```{r,include=F,echo=F,eval=FALSE}
mtx_dir <- "/Users/mattpaul/Downloads/filtered_feature_bc_matrix/"

#/10k_PBMC_3p_nextgem_Chromium_Controller_filtered_feature_bc_matrix.tar.gz"
```

```{r load_data,include=TRUE,eval=FALSE}
library(Seurat)
#
mtx_dir <- "path to matrix directory"
mtx <- Seurat::Read10X(mtx_dir)
#
is(mtx)
head(mtx)
```

## Load cellranger matrix (in .h5 format) into matrix
- In an alternative way, we can use *Read10X_h5* function to read cellranger matrix from an **.h5** file.
- What is difference? Why would do it differently?


```{r load_h5,include=TRUE,eval=FALSE}
h5_file <- "path to matrix h5 file"
h5_file <- "~/Downloads//10k_PBMC_3p_nextgem_Chromium_Controller_molecule_info.h5"
mtx <- Seurat::Read10X_h5(h5_file)
#
mtx
```

## Create Seurat object from matrix - defaults for this are 0? Why different? Should include?
* The matrix is loaded into Seurat object with **CreateSeurtObject()**
* Cut-off
  + min_gene: minimal genes detected in each cell. *(cells with too few genes detected)*
  + min_cell: minimal cells a given gene was tested positive. *(Remove genes expressed in too few cells)*
```{r load_CreateOBJ,include=TRUE,eval=FALSE}
sample_id <- "PBMC_10k" # sample name
min_gene <- 200 # cut-off for cells with too few genes detected
min_cell <- 10 # cut-off to remove genes expressed in too few cells
#
seu_obj <- Seurat::CreateSeuratObject(mtx,project=sample_id,min.cells=min_cell,min.features=min_gene)
seu_obj[["dset"]] <- sample_id # Create a category for sample
seu_obj <- Seurat::RenameCells(seu_obj,add.cell.id=sample_id) # add sample name in front of cell barcode
```

```{r laod_CreatOBJ_pres,include=TRUE,eval=TRUE}
message("Seurat Object")
seu_obj
#
message("Metadata information")
head(seu_obj,2)
```

## Estimate proportions of mitochondrial genes - what is bad proporiotn?
2 methods. do we need to cover both?
* Estimate the counts of mitochondrial genes in overall counts with **PercentageFeatureSet()**
* The Feature sets can be selected by *patterns of gene names* or *directly assign specific genes*. 
  + While assign gene names, we have to make sure *the given genes are detected in the Seurat object*.
```{r load_estMT,include=TRUE,eval=FALSE}
message("select by pattern")
seu_obj[["percent.mt"]] <- PercentageFeatureSet(seu_obj,pattern = "^MT-")
summary(seu_obj$percent.mt)
#
message("assign gene names")
mt_gene <- c("MT-ND1","MT-ND2","MT-ND3","MT-ND4","MT-ND4L","MT-ND5","MT-ND6",
             "MT-CO1","MT-CO2","MT-CO3","MT-ATP6","MT-ATP8","MT-CYB")
mt_gene_det <- mt_gene[mt_gene %in% rownames(seu_obj)] # make sure the genes presented in the seurat object
seu_obj[["percent.mt2"]] <- PercentageFeatureSet(seu_obj,features = mt_gene_det)
summary(seu_obj$percent.mt2)
#
message("Test condordance with Spearman's correlation")
dat <- data.frame(byPattern=seu_obj$percent.mt,byGene=seu_obj$percent.mt2,stringsAsFactors = FALSE)
cor_val <- cor.test(dat$byPattern,dat$byGene,method = "spearman")
ggplot(dat,aes(x=byPattern,y=byGene))+geom_point()+geom_smooth()+
  labs(x="% of MT,est by pattern",y="% of MT, est by genes",
       subtitle = paste0("rho=",round(cor_val$estimate,3),"; p-value=",cor_val$p.value[1]))+
  theme_classic()
```

# Normalization, Feature selection, and data scaling
- Log normalization
- SCTransform

## Log normlaization ~ default 
* In Seurat vignette, this step includes:
  + log normalization with *NomalizeData()*
  + Identify Variable Features with *FindVariableFeatures()*
  + Scale Data with *ScaleData()*
```{r norm_log,include=TRUE,eval=FALSE}
message("log normalization, scale to 10,000")
seu_obj <- NormalizeData(seu_obj,normalization.method="LogNormalize",scale.factor=10000)
#
message("Identify variables, select top 3000 features with VST method")
seu_obj <- FindVariableFeatures(seu_obj,select.method="vst",nfeatures=3000)
#
message("Scale data")
seu_obj <- ScaleData(seu_obj)
```

## Highly Variable Features
- Highly variable features were colored with *red*. 
- Gene symbols of the top 10 higly variable features were labeled near by the spots
```{r norm_plotHVF,include=TRUE,eval=TRUE}
top10 <- head(VariableFeatures(seu_obj), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(seu_obj)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

## Normalization and scaling by usign SCTransform
- SCTransform is a normalization based on negative binomial regression
- It is performed with **SCTransform()** function of Seurat.
- This function equal to the combinations of NormalizaeData(), FindVariableFeatures(), and ScaleData().
- The normalized data is stoered in assay **SCT**.
```{r norm_sct,include=TRUE,eval=FALSE}
seu_obj <- SCTransform(seu_obj,variable.features.n = 3000)
#
seu_obj
```

## Compare the results between log normalization and SCTransform

- pros and cons?
- do we have any plots to describe this?

```{r norm_comp,include=TRUE,eval=FALSE}
message("get log normalization data")
log_mat <- GetAssayData(seu_obj,assay="RNA",slot="data")
log_mat <- as.matrix(log_mat)
log_avgExp <- rowMeans(log_mat)
#
message("get SCTransformed data")
sct_mat <- GetAssayData(seu_obj,assay="SCT",slot="data")
sct_mat <- as.matrix(sct_mat)
sct_avgExp <- rowMeans(sct_mat)
#
message("Test concordance with Spearman's correlation")
dat <- data.frame(logNorm=log_avgExp,SCT=sct_avgExp,stringsAsFactors = FALSE)
cor_val <- cor.test(log_avgExp,sct_avgExp,method = "spearman")
ggplot(dat,aes(x=logNorm,y=SCT))+geom_point()+geom_smooth()+
  labs(x="Log_Normalization",y="SCTransform",subtitle = paste0("rho=",round(cor_val$estimate,3),"; p-value=",cor_val$p.value[1]))+
  theme_classic()
```

# Cell cycle Phases
In this step, we will introduce two methods to estimate cell cycle phase for each cell.
- The CellCycleScoring function in *Seurat*
- Cyclone function in *scran* 

## Estimate with CellCycleScoring function of Seurat
* Requirement
  + The Seurat object after normalization
  + Features for S-Phase (feat_s)
  + Features for G2M-Phase (feat_g2m)
  + Seurat provide features for S/G2M-Phase
```{r ccPhase_Seurat,include=TRUE,eval=FALSE}
message("Features for S-Phase")
feat_s <- cc.genes$s.genes
feat_s
#
message("Features for G2M-Phase")
feat_g2m <- cc.genes$g2m.genes
feat_g2m
```

## Estimate with CellCycleScoring function of Seurat
* Rationale
  + Estimate the scoring based on specific features for S-phase/G2M-phase, respectively
  + Differentiate Phase by the scores.
  + For a given cell with significant high S.Score or G2M.Score was assigned as S/G2M, respectively
  + Cells with low both S.Score and G2M.Score were assigned as G1.
* Would *overestimate* cells in S-/G2M-phases in the tissues with low cell cycle, eg neurons (what does this look like?)
```{r ccPhase_plot_Seurat,include=TRUE,eval=FALSE}
seu_obj <- CellCycleScoring(seu_obj,s.features = feat_s,g2m.features = feat_g2m)
#
dat_s <- data.frame(cell_id=Cells(seu_obj),cat="S_Score",Phase=seu_obj$Phase,score=seu_obj$S.Score,stringsAsFactors = FALSE)
dat_g2m <- data.frame(cell_id=Cells(seu_obj),cat="G2M_Score",Phase=seu_obj$Phase,score=seu_obj$G2M.Score,stringsAsFactors = FALSE)
dat <- rbind(dat_s,dat_g2m)
#
dat$Phase <- factor(dat$Phase,levels = c("G1","S","G2M"))
ggplot(dat,aes(x=Phase,y=score,fill=Phase))+geom_boxplot()+
  labs(x="",y="Score",fill="Phase")+
  facet_wrap(~cat)+theme_classic()
```


## Cell cycle estimation by cyclone
* This step is processed by cyclon function in scran
* Requirement
+ convert Seurat object to SinlgeCellExperiment Object
+ features for G1, S, and G2M Phases
* Rationale
+ Estimate scores for each cell in different phases with the corresponding features
+ Assign the cell cycle phases by the maximum of scores
* Much *slower* than CellCycleScoring in Seurat

## Requirement for cyclon

how do we make that object?
```{r ccPhase_cyclon_prep,include=TRUE,eval=FALSE}
library(scran)
#
message("convert seurat object into SingleCellExperiment")
sce <- as.SingleCellExperiment(seu_obj,assay = "RNA")
rowData(sce)$SYMBOL <- rownames(sce)
sce
#
message("Features for cell cycle phases")
load("~/Downloads/ccGene_mouse_human_geneSymbol_ensemblID_20220817.RData")
ccGene_hs <- ccGene_mm_hs$human_symbol
lapply(ccGene_hs,function(x){head(x,2)})
```

## Process with cyclone
```{r ccPhase_cyclon_proc,include=TRUE,eval=FALSE}
assignments <- cyclone(sce, ccGene_hs, gene.names=rowData(sce)$SYMBOL)
lapply(assignments,head)
#
seu_obj[["cyclon_Phase"]] <- assignments$phases
seu_obj[["cyclon_G1Score"]] <- assignments$scores$G1
seu_obj[["cyclon_SScore"]] <- assignments$scores$S
seu_obj[["cyclon_G2MScore"]] <- assignments$scores$G2M
```

## Score vs Phases
```{r ccPhase_cyclon_boxPlot,include=TRUE,eval=FALSE}
dat_g1 <- data.frame(cell_id=Cells(seu_obj),cat="cyclon_G1Score",Phase=seu_obj$cyclon_Phase,score=seu_obj$cyclon_G1Score,stringsAsFactors = FALSE)
dat_s <- data.frame(cell_id=Cells(seu_obj),cat="cyclon_SScore",Phase=seu_obj$cyclon_Phase,score=seu_obj$cyclon_SScore,stringsAsFactors = FALSE)
dat_g2m <- data.frame(cell_id=Cells(seu_obj),cat="cyclon_G2MScore",Phase=seu_obj$cyclon_Phase,score=seu_obj$cyclon_G2MScore,stringsAsFactors = FALSE)
dat <- rbind(dat_g1,dat_s,dat_g2m)
#
dat$Phase <- factor(dat$Phase,levels = c("G1","S","G2M"))
dat$cat <- factor(dat$cat,levels = c("cyclon_G1Score","cyclon_SScore","cyclon_G2MScore"))
ggplot(dat,aes(x=Phase,y=score,fill=Phase))+geom_boxplot()+
  labs(x="",y="Score",fill="Phase")+
  facet_wrap(~cat)+theme_classic()
```

## Compare the two strategies
- *cyclone* assigned more cells to *G1* phase. (*more accurate?*). seems like a very big discrepancy?
- *cyclone* spends more time than *CellCycleScoring*.
- It might be worth to process both methods and evaluate which results make more sense for your own dataset.
```{r ccPhase_comp,include=TRUE,eval=FALSE}
message("Cell Cycle Phase determined by Seurat::CellCycleScoring()")
table(seu_obj$Phase)
#
message("Cell Cycle Phase determined by scran::cyclone()")
table(seu_obj$cyclon_Phase)
#
message("Compare the two method: Seurat in row and scran in column")
table(seu_obj$Phase,seu_obj$cyclon_Phase)
```

# Detect doublets with Scrublet

## Description
* Doublets mean multiple cells clumped in the same single droplet. 
+ A doublet may have much more UMIs and Genes in a single cell barcode than the overall population
+ Multiple marker genes, even they are exclusive, can be detected in doublets at the same time. 

* Scrublet is a tool to detect doublets
+ The GitHub [link](https://github.com/swolock/scrublet) 
+ Original Paper on bioRxiv [link](https://www.biorxiv.org/content/10.1101/357368v1)
+ It's based on python. So, we have to use reticulate to invoke scrublet into R.

## Estimate doublet score
```{r}
library(Herper)

conda_install  <- install_CondaTools("scrublet","scRNA",pathToMiniConda = "../mini")

Sys.setenv('RETICULATE_PYTHON'=file.path(conda_install$pathToEnvBin, "python"))
```


```{r det_doublet_est,include=TRUE,eval=FALSE}


message("invoke scrublet in R by using reticulate")
# library(reticulate)
# py_path <- "path to the python with scrublet installed"
# # py_path <- "/Users/JDLuo/opt/anaconda3/envs/reticulate/bin/python"
# Sys.setenv('RETICULATE_PYTHON'=py_path)
scr <- import("scrublet")
#
message("dumping count matrix from seurat object")
mat <- GetAssayData(seu_obj,assay = "RNA",slot = "counts")
mat <- as.matrix(mat)
#
message("Estimate doublet score")
scrub <- scr$Scrublet(t(mat))
doublet <- scrub$scrub_doublets()
names(doublet) <- c("doublet_score","doublet")
#
message("doublet score")
summary(doublet$doublet_score)
#
message("doublets")
table(doublet$doublet)
```

## Import Scrublet results into Seurat object
```{r det_doublet_pres,include=TRUE,eval=FALSE}
seu_obj[["doublet_score"]] <- doublet$doublet_score
seu_obj[["doublet"]] <- doublet$doublet
#
message("UMIs,genes, and doublet scores of each cell in doublets/singlets")
VlnPlot(seu_obj,group.by = "doublet",
        features = c("nCount_RNA","nFeature_RNA","doublet_score"),
        pt.size = 0)
#
FeatureScatter(seu_obj,feature1 = "nCount_RNA",feature2 = "nFeature_RNA",pt.size = 0.1,group.by = "doublet")
```

# Quality assessment

## Distribution of particular variables in violin plot
* This kind of plot is applied to evaluate the variables, including
  + UMIs (*nCount_RNA*) per cell
  + Genes detected (*nFeature_RNA*) per cell
  + Ratio of UMIs of mitochondrial genes to nucleus genes (*percent.mt*)
* If a bimodal distribution is identified, we need to check any interference

nFeature is bimodal
```{r qcPlot_vlnPlot_pres1,include=TRUE,eval=FALSE}
VlnPlot(seu_obj,group.by = "dset",
        features = c("nCount_RNA","nFeature_RNA","percent.mt"),
        pt.size = 0)
```

## Evaluate the interactions between two variables with scatter plots
* Comparison between UMIs and genes detected per cell
  + UMIs (*nCount_RNA*) and the genes detected (*nFeature_RNA*) shall be in a high correlation
  + doublets would be with high UMIs and high genes detected.
```{r qcPlot_scatter_pres1,include=TRUE,eval=FALSE}
message("UMIs vs genes detected per cell, group by doublet")
FeatureScatter(seu_obj,feature1 = "nCount_RNA",feature2 = "nFeature_RNA",
               group.by = "doublet")
```

* Comparison between UMI counts and the percent of mitochondrial content
  + Potential cell debris would show low UMI counts (*nCount_RNA*) and high percentage of mitochondrial genes (*percent.mt*)
```{r qcPlot_scatter_pres2,include=TRUE,eval=FALSE}
message("UMIs vs the ratio of mitochondrial genes")
FeatureScatter(seu_obj,feature1 = "nCount_RNA",feature2 = "percent.mt")
```

## Ridge plot
* This plot would be applied to demonstrate a given variable and their corresponding score (eg. doublet vs doublet_score).
* It is applied to evaluate the difference between groups in the variable.
* This plot would be widely applied in hash-tag determination of CITE-Seq. We could discuss more details in Section III.
```{r qcPlot_ridgePlot_pres,include=TRUE,eval=FALSE}
message("doublet vs doublet_score")
RidgePlot(seu_obj,group.by = "doublet",features = c("doublet_score"))
```

# Filtering doublets and cell debris - why does debris have high mt?

## Criteria
* The cell debris are with high percent.mt and low UMI counts. Generally, we set various cut-off on percent.mt
  + In most scRNA cases, we set the *percent.mt > 10* (95% of overall population)
  + For several specific tissues with high oxygen consumption, like activated leukocytes or muscles, we set the *percent.mt > 25*.
  + For single-nuclei profiling, we should not get any UMIs originated from mitochondrial genes. We set the *percent.mt > 1*.
* The doublets detected by Scrublet shall be removed.
```{r filtCell_pres,include=TRUE,eval=FALSE}
message("cells shall be moved")
table(seu_obj$doublet=="TRUE" | seu_obj$percent.mt >= 10)
message("filtering cells")
seu_filt <- subset(seu_obj,subset=doublet=="FALSE" & 
                     percent.mt < 10)
seu_filt
```

# Evaluate the percent of variances. can explained by particular variables - is this a per cell value?
* This step is applied to identify the unexpected variables contribute to a high percent of the variances. (Also known as **confounders**)
* Once the confounders are identified, we have to minimize the effectiveness by using **SCTranform()** or **ScaleData()**.
* The functions *getVarianceExplained* and *plotExplanatoryVariables* in *scater* are applied.

```{r evalVarExp,include=TRUE,eval=FALSE}
library(scater)
sce <- as.SingleCellExperiment(seu_filt,assay = "RNA")
vars <- getVarianceExplained(sce, 
                             variables = c("percent.mt","nCount_RNA","nFeature_RNA",
                                           "doublet_score",
                                           "cyclon_G1Score","cyclon_SScore","cyclon_G2MScore","cyclon_Phase"))
plotExplanatoryVariables(vars)
```

```{r scaleVar,include=TRUE,eval=FALSE}
seu_filt <- SCTransform(seu_filt,vars.to.regress = c("nCount_RNA","nFeature_RNA","cyclon_SScore"))
#
sce <- as.SingleCellExperiment(seu_filt,assay = "SCT")
vars <- getVarianceExplained(sce, 
                             variables = c("percent.mt","nCount_RNA","nFeature_RNA",
                                           "doublet_score",
                                           "cyclon_G1Score","cyclon_SScore","cyclon_G2MScore","cyclon_Phase"))
plotExplanatoryVariables(vars)
```

# Dimension Reduction
- Perform linear dimensional reduction with Principle Component Analysis (PCA)
- Determine dimensionality with Elbow plot
- Process non-linear dimensional reduction with UMAP
- to select major PCs for the following clustering 

## Linear dimensional reduction
Perform PCA with *RunPCA()* function in Seurat. Set Principle components to 30.

```{r dimRed_pca,include=TRUE,eval=FALSE}
set.seed(1001)
seu_filt <- SCTransform(seu_filt)
seu_filt <- RunPCA(seu_filt,assay = "SCT",npcs = 30)
```

## Evaluate dimensionality with elbow plot
```{r dimRed_elbow,include=TRUE,eval=TRUE}
ElbowPlot(seu_filt,ndims = 30,reduction = "pca")
```

## Elbow plot with color

create cutoff based on if variabilty is low, and you already have most of it, 
or if variabiltiy is alrgely not changing. 
```{r dimRed_elbow2,include=TRUE,eval=FALSE}
library(dplyr)
# Determine percent of variation associated with each PC
pct <- seu_filt[["pca"]]@stdev / sum(seu_filt[["pca"]]@stdev) * 100 
# Calculate cumulative percents for each PC
cumu <- cumsum(pct) 
# Determine which PC exhibits cumulative percent greater than 90% and % variation associated with the PC as less than 5
co1 <- which(cumu > 90 & pct < 5)[1] 
# Determine the difference between variation of PC and subsequent PC, last point where change of % of variation is more than 0.1%.
co2 <- sort(which((pct[1:length(pct) - 1] - pct[2:length(pct)]) > 0.1), decreasing = T)[1] + 1 
pcs <- min(co1, co2)
# Create a dataframe with values
plot_df <- data.frame(pct = pct,cumu = cumu,rank = 1:length(pct))
```

```{r dimRed_elbow2_rep,include=TRUE,eval=FALSE}
message("Elbow plot to visualize")
ggplot(plot_df, aes(cumu, pct, label = rank, color = rank > pcs)) + geom_text() + 
  geom_vline(xintercept = 90, color = "grey") + 
  geom_hline(yintercept = min(pct[pct > 5]), color = "grey") +
  theme_bw()
#
message("PCs used")
pc <- pcs
pc
```

## Make non-linear dimensional reduction - why UMAP? What is it? Explain,
* Estimate neighbors with **FindNeighors()** function in *Seurat*
  + reduction method: *PCA*
  + use dimessions from 1 to the PCs detected by elbow plot
* Make non-linear dimensional reduction with UMAP
```{r dimRed_UMAP,include=TRUE,eval=FALSE}
seu_filt <- FindNeighbors(seu_filt,dims = 1:pc,reduction = "pca")
seu_filt <- RunUMAP(seu_filt,dims = 1:pc,reduction = "pca")
```

# Clustering
- The cells are clustered based on UMAP by using *FindClusters()* function in Seurat.
- The clustering is differentiated by **original Louvain algorithm**.
- To find out a better resolution, we test clustering with different resolutions and evluate in a clustree

## Identify clusters with default Louvain algorithm
```{r clust_default,include=TRUE,eval=FALSE}
seu_filt <- FindClusters(seu_filt,resolution = 0.5)
seu_filt[["cluster_byDefault"]] <- seu_filt$seurat_clusters
DimPlot(seu_filt,group.by = "seurat_clusters",label = TRUE,pt.size = 0.2)+NoLegend()
```

## Test resolution from 0.1 to 1.0
```{r clust_resoEval,include=TRUE,eval=FALSE}
library(clustree)
library(ggplot2)
#
# resolution from 0.1 to 1.0
reso <- c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)
reso_res <- lapply(1:length(reso),function(x,seu_filt,reso){
  seu_filt <- FindClusters(seu_filt,resolution = reso[x])
  clust <- setNames(seu_filt$seurat_clusters,Cells(seu_filt))
  return(clust)},seu_filt,reso)
names(reso_res) <- paste0("k",1:length(reso))
k_tab <- do.call(cbind,reso_res)
k_dat <- as.data.frame(k_tab)
```

## Evaluate resolutions with clustree
- The resolution from top to the bottom are *0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1*
- To identify the better resolution
  + less cross talks between lineages
  + more clusters identified
* In this case, we use *resolution=0.7*
```{r clust_resoEval2,include=TRUE,eval=FALSE}
clustree(k_dat, prefix = "k", node_colour = "sc3_stability")
```

## Clustering
```{r clust_resoFix,include=TRUE,eval=FALSE}
seu_filt <- FindClusters(seu_filt,resolution = 0.2)
DimPlot(seu_filt,group.by = "seurat_clusters",label = TRUE,pt.size = 0.2)+NoLegend()
```

# Evaluate Results
## Test with know markers - where are these markers fom?
```{r resEval_knownMarker,include=TRUE,eval=FALSE}
FeaturePlot(seu_filt, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A", "FCGR3A", "LYZ", "PPBP",
    "CD8A","IL7R","CCR7","S100A4","CST3"))

message("Naive CD4 T-cells")
VlnPlot(seu_filt,group.by = "seurat_clusters",features = c("IL7R","CCR7"))
#
message("CD14+ Mono")
VlnPlot(seu_filt,group.by = "seurat_clusters",features = c("CD14","LYZ"))
#
message("Memory CD4 T-cells")
VlnPlot(seu_filt,group.by = "seurat_clusters",features = c("IL7R","S100A4"))
#
message("B-cells")
VlnPlot(seu_filt,group.by = "seurat_clusters",features = c("MS4A1"))
#
message("CD8_ T-cells")
VlnPlot(seu_filt,group.by = "seurat_clusters",features = c("CD8A"))
#
message("FCGR3A+ Mono")
VlnPlot(seu_filt,group.by = "seurat_clusters",features = c("FCGR3A","MS4A7"))
#
message("NK")
VlnPlot(seu_filt,group.by = "seurat_clusters",features = c("GNLY","NKG7"))
#
message("DC")
VlnPlot(seu_filt,group.by = "seurat_clusters",features = c("FCER1A","CST3"))
#
message("Platelet")
VlnPlot(seu_filt,group.by = "seurat_clusters",features = c("PPBP"))
#
```

```{r }
sce <- as.SingleCellExperiment(seu_filt,assay = "RNA")
vars <- getVarianceExplained(sce, 
                             variables = c("percent.mt","nCount_RNA","nFeature_RNA",
                                           "doublet_score","seurat_clusters",
                                           "cyclon_G1Score","cyclon_SScore","cyclon_G2MScore","cyclon_Phase"))
plotExplanatoryVariables(vars)
```


