
---
## Customizing your differentials
Seurat has many different options for differentials. You can use these to fully customize your analysis.

- Assign the groups we want to compare: *group_by*
- If we want to get a unfiltered results: set *logfc.threshold = 0* and *min.pct = 0*
- Seurat also support different statistic methods: *Wilcox Rank Sum test (wilcox)*, *likelihood-ratio test (bimod)*,*receiver operating characteristic (roc)*,*Student's T-test (t)*,*negative binomial linear model (negbinom)*, *poisson generalized linear model (poisson)*, *logitstic regression (LR)*, *MAST*, *DESeq2*


---
## Export clusters and annotations

```{r}
meta <- seu_filt@meta.data
clust_dat <- data.frame(Barcode=Cells(seu_filt),cluster=meta$seurat_clusters)
clust_dat$Barcode <- gsub("PBMC_10k_","",clust_dat$Barcode)
head(clust_dat)

write.table(clust_dat,file = "clust_dat.csv",
            sep=",",quote = FALSE,row.names = FALSE)

cellType_dat <- data.frame(Barcode=Cells(seu_filt),cellType=meta$cellType_byClust)
cellType_dat$Barcode <- gsub("PBMC_10k_","",cellType_dat$Barcode)
head(cellType_dat)

write.table(cellType_dat,file = "cellType_dat.csv",
            sep=",",quote = FALSE,row.names = FALSE)
```

---
## Export Seurat results into Cloupe file
We can export the UMAP and annotations of Seurat object and import them into cloupe file.
* Embedding is applied to extract dimension reduction (eg. PCA, UMAP, tSNE) into matrix
  + This data frame include: Barcode, X (UMAP_1), and Y (UMAP_2)
  + Need to match Barcode to the barcode in cloupe file
  + Store in a csv format

```{r}
umap <- Embeddings(seu_filt,reduction = "umap")
head(umap,2)

umap_tab <- data.frame(Barcode=rownames(umap),
                       "X coordinate"=umap[,1],Y=umap[,2])
umap_tab$Barcode <- gsub("PBMC_10k_","",umap_tab$Barcode)
head(umap_tab,2)

write.table(umap_tab,file = "umap_dat.csv",
            sep=",",quote = FALSE,row.names = FALSE)
```


---
## Export clusters and annotations

```{r}
meta <- seu_filt@meta.data
clust_dat <- data.frame(Barcode=Cells(seu_filt),cluster=meta$seurat_clusters)
clust_dat$Barcode <- gsub("PBMC_10k_","",clust_dat$Barcode)
head(clust_dat)

write.table(clust_dat,file = "clust_dat.csv",
            sep=",",quote = FALSE,row.names = FALSE)

cellType_dat <- data.frame(Barcode=Cells(seu_filt),cellType=meta$cellType_byClust)
cellType_dat$Barcode <- gsub("PBMC_10k_","",cellType_dat$Barcode)
head(cellType_dat)

write.table(cellType_dat,file = "cellType_dat.csv",
            sep=",",quote = FALSE,row.names = FALSE)
```




